Exploring the Composition of Lithic Assemblages in Mesolithic South-Eastern Norway

### 2022-01-04 analysis/paper/paper.Rmd
Edits after review

---
title: "Exploring the Composition of Lithic Assemblages in Mesolithic South-Eastern Norway"
header-includes:
   - \usepackage{lineno}
   - \linenumbers
author:
  - Isak Roalkvam:
      email: isak.roalkvam@iakh.uio.no
      institute: [UiO]
      correspondence: false
institute:
  - UiO: University of Oslo, Department of Archaeology, Conservation and History
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      toc: false
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/saa.csl" # Insert path for the bib-style
abstract: |
  This paper leverages multivariate statistics to explore the composition of 54 Mesolithic assemblages located in south-eastern Norway. To provide analytical control pertaining to factors such as variable excavation practices, systems for artefact categorisation and raw-material availability, the sites chosen for analysis have all been excavated relatively recently and have a constrained geographical distribution. The assemblages were explored following two strains of analysis. The first of these entailed the use of artefact categories that are established within Norwegian Mesolithic archaeology, while the other involved drawing on measures that have been linked directly to land-use and mobility patterns associated with lithic assemblages more widely. The findings pertaining to the established artefact categories largely reflect the temporal development previously reported in Norwegian Mesolithic research, which has been based on more subjectively driven methods. Furthermore, the chronological trends associated with variables taken from the so-called Whole Assemblage Behavioural Indicators (e.g. Clark and Barton 2017), originally devised for characterising Palaeolithic assemblages in terms of associated mobility patterns, also align with the development previously proposed in the literature. This provides an initial indication that these measures are applicable in a Norwegian Mesolithic setting as well, setting the stage for a more targeted and rigorous model evaluation outside this exploratory setting. Furthermore, this finding supports the notion that these measures can offer a powerful comparative tool in the analysis of lithic assemblages more generally.

@@

A defining characteristic of the Norwegian Mesolithic is that a clear majority of the known sites are located in coastal areas [e.g. @bjerck2008]. Furthermore, these coastal sites appear to predominantly have been located on or close to the contemporary shoreline when they were in use [@åstveit2018; @breivik2018; @møller1987; @solheim2020]. In south-eastern Norway, this pattern is combined with a continuous regression of the shoreline, following from isostatic rebound [e.g. @romundset2018; @sørensen1979]. The fairly rapid shoreline displacement means that the sites tend not to have retained their strategic or ecologically beneficial shore-bound location for long periods of time [cf. @perreault2019, 47]. Consequently, the shore-bound settlement, combined with the rapid shoreline displacement has resulted in a relatively high degree of spatial separation of cumulative palimpsests, to follow the terminology of @bailey2007, while the reconstruction of the trajectory of relative sea-level change allows for a relatively good control of when these accumulation events occurred. In other parts of the world, a higher degree of spatial distribution means that while the physical separation of material can help delineate discrete events, this typically comes at the cost of losing temporal resolution as any stratigraphic relationship between the events is lost [@bailey2007].

@@

The lithic data analysed are based on the classification of the site assemblages done for the original excavation reports, and consists of `r ncol(select(sites, core:flake_retouch_nonflint))` variables representing different debitage and tool types. The artefact data have been divided into flint and non-flint materials. Flint does not outcrop naturally in southern Norway, and is only available locally as nodules that have been transported and deposited by retreating and drifting ice [e.g. @berg-hansen1999]. This means that the distribution and quality of flint has been impacted by a diverse set of climatic and geographical factors [@eigeland2015, 46]. Thus, while flint is treated as a unified category here, the variability in quality could have been substantial. Furthermore, the various non-flint raw materials that have been lumped together have quite disparate properties, where fine-grained cryptocrystalline materials are often used as a substitute or supplement to flint, while other, coarser materials are usually associated with the production of axes and other macro tools. Given this differentiated use, these raw-material properties are expected to be reflected in the retained debitage and tool categories. An important benefit of combining all of the non-flint materials is that this reduces the dependency on whether or not these have been correctly and consistently categorised for the reports [cf. @frivoll2017]. Finally, while factors such as landscape changes through shoreline displacement can have led to variable raw-material availability at the analysed sites, their relatively constrained geographical distribution hopefully counteracts some non-behavioural sources of variation.

@@

```{r variable-aggregation, fig.cap = "Aggregation of variables for the correspondence analysis. The column on the left shows the data as originally compiled. The column on the right shows how these have been aggregated for the analysis.", fig.height = 7}
# Most of the code for assembling the Sankey plot has been taken directly from 
# various walkthroughs at https://www.r-graph-gallery.com

@@

# Create plot
sn <- sankeyNetwork(Links = links, Nodes = nodes,
              Source = "IDsource", Target = "IDtarget",
              NodeID = "name", Value = "value", colourScale = my_colour,
              sinksRight = FALSE, LinkGroup = "new_categories", 
              NodeGroup="group", nodeWidth = 12, nodePadding = 7, fontSize = 12)

@@

```{r cor-plot, fig.cap = "Correlation matrix showing transformation of skewed variables for the PCA. The mean age of the sites has also been included to visualise overall temporal trends. Cells below the diagonal display the bivariate distributions with a fitted OLS-regression. The cells above the diagonal display and are coloured by the corresponding Pearson's correlation coefficient.", fig.height = 9.5, fig.width = 10}
# Transform the skewed variables - sqrt() is used where there are zero values,
# and seems to provide an adequate result.
lithicstr <- lithics2 %>%
  select(-site) %>% 
  # Transform skewed variables and reorder
  mutate(
         "log10(lithic_dens)" = log10(lithic_dens),
         "log10(secondary_prop)" = log10(secondary_prop),
         "core_prop^(1/3)" = core_prop^(1/3),
         "nonflint_prop^(1/3)" = nonflint_prop^(1/3), .keep = "unused") %>% 
  relocate("mean_date",
    "log10(lithic_dens)", "nonflint_prop^(1/3)", "blanks_prop",
    "log10(secondary_prop)", "chip_prop", "core_prop^(1/3)" 
  )

@@

The flint material of the later sites are to larger extent characterised by micro-blades, which corresponds to the transition to micro-blade production from handle cores [e.g. @solheim2020]. A more fragmented flint material, as indicated by the relative importance of flint chips and fragments, is also a previously noted property of some later Mesolithic, as well as early Neolithic sites [e.g. @fossum2017; @stokke2018a]. The most defining material for the later sites, however, is non-flint macro tools and associated debitage, which is dominating some of these assemblages. It was noted above that this material does not seem to impact the latest sites, which would indicate that specialised axe production sites disappear towards the end of the Mesolithic, a notion that would be in line with previous suggestions [e.g. @glørstad2011; @reitan2016].

@@

Supplementary material

@@

splt1 <- fviz_ca_row(nonprop, col.row = "black",
            fill = all_nonflint_prop$mean_date / 1000, title = "A",
                     repel = TRUE, shape = 21, pointsize = 2.9) +
  scale_fill_gradient2(name = "Mean age kyr BCE", low="yellow", mid="orange", 
                        high="red", midpoint = centdate) +
  theme(legend.position = "bottom")

@@

```{r, fig.height = 15, fig.cap = "Same as above, only that here the proportion of non-flint is used as a supplementary column. The negligble difference between this CA and that above indicates that the raw material variability is integrated in the different artefact types, and therefore that the effect of artefact types and raw material cannot be seperated. These plots thus hide important variability that is captured in the ones presented in the main text. A) Row plot, B) Column plot."}

# Version of the above where the proportion of non-flint is simply added as a 
# supplementary column, followng Reviewer 2. 
moddat <- select(all_nonflint_prop, -site, -mean_date, -flint_count, 
                       -reported_start_bc, -reported_end_bc,
                       -askeladden_id)

@@

splt4 <- fviz_ca_col(nonpropsup, col.col.sup = "black", col.col = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE,
             title = "B") +
  labs(colour = "Contribution to total inertia") +
  theme(legend.position = "bottom")

>>>>

---
title: "Exploring the Composition of Lithic Assemblages in Mesolithic South-Eastern Norway"
header-includes:
   - \usepackage{lineno}
   - \usepackage{afterpage}
   - \linenumbers
author:
  - Isak Roalkvam:
      email: isak.roalkvam@iakh.uio.no
      institute: [UiO]
      correspondence: false
institute:
  - UiO: University of Oslo, Department of Archaeology, Conservation and History
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      toc: false
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/saa.csl" # Insert path for the bib-style
abstract: |
  This paper leverages multivariate statistics to explore the composition of 54 Mesolithic assemblages located in south-eastern Norway. To provide analytical control pertaining to factors such as variable excavation practices, systems for artefact categorisation and raw-material availability, the sites chosen for analysis have all been excavated relatively recently and have a constrained geographical distribution. The assemblages were explored following two strains of analysis. The first of these entailed the use of artefact categories that are in established use within Norwegian Mesolithic archaeology, while the other involved drawing on measures that have been linked directly to land-use and mobility patterns associated with lithic assemblages more widely. The findings pertaining to the established artefact categories largely reflect the temporal development previously reported in Norwegian Mesolithic research, which has been based on more subjectively driven methods. Furthermore, the chronological trends associated with variables taken from the so-called Whole Assemblage Behavioural Indicators (e.g. Clark and Barton 2017), originally devised for characterising Palaeolithic assemblages in terms of associated mobility patterns, also align with the development previously proposed in the literature. This provides an initial indication that these measures are applicable in a Norwegian Mesolithic setting as well, setting the stage for a more targeted and rigorous model evaluation outside this exploratory setting. Furthermore, this finding supports the notion that these measures can offer a powerful comparative tool in the analysis of lithic assemblages more generally.

@@

A defining characteristic of the Norwegian Mesolithic is that a clear majority of the known sites are located in coastal areas [e.g. @bjerck2008]. Furthermore, these coastal sites appear to predominantly have been located on or close to the contemporary shoreline when they were in use [@åstveit2018; @breivik2018; @møller1987; @solheim2020a]. In south-eastern Norway, this pattern is combined with a continuous regression of the shoreline, following from isostatic rebound [e.g. @romundset2018; @sørensen1979]. The fairly rapid shoreline displacement means that the sites tend not to have retained their strategic or ecologically beneficial shore-bound location for long periods of time [cf. @perreault2019, 47]. Consequently, the shore-bound settlement, combined with the rapid shoreline displacement has resulted in a relatively high degree of spatial separation of cumulative palimpsests, to follow the terminology of @bailey2007, while the reconstruction of the trajectory of relative sea-level change allows for a relatively good control of when these accumulation events occurred. In other parts of the world, a higher degree of spatial distribution means that while the physical separation of material can help delineate discrete events, this typically comes at the cost of losing temporal resolution as any stratigraphic relationship between the events is lost [@bailey2007].

@@

The lithic data analysed are based on the classification of the site assemblages done for the original excavation reports, and consists of `r ncol(select(sites, core:flake_retouch_nonflint))` variables representing different debitage and tool types. The artefact data have been divided into flint and non-flint materials. Flint does not outcrop naturally in southern Norway, and is only available locally as nodules that have been transported and deposited by retreating and drifting ice [e.g. @berg-hansen1999]. This means that the distribution and quality of flint has been impacted by a diverse set of climatic and geographical factors [@eigeland2015, 46]. Thus, while flint is treated as a unified category here, the variability in quality could have been substantial. Furthermore, the various non-flint raw materials that have been lumped together have quite disparate properties, where fine-grained cryptocrystalline materials are often used as a substitute or supplement to flint, while other, coarser materials are usually associated with the production of axes and other macro tools. Given this differentiated use, these raw-material properties are expected to be reflected in the retained debitage and tool categories. An important benefit of combining all of the non-flint materials is that this reduces the dependency on whether or not these have been correctly and consistently categorised for the reports [cf. @frivoll2017]. Finally, while factors such as landscape changes through shoreline displacement can have led to variable raw-material availability at the analysed sites, for example by impacting accessibility by means of watercraft, the relatively constrained geographical distribution of the sites hopefully counteracts some environmentally given sources of variation.

@@

```{r variable-aggregation, fig.cap = "Aggregation of variables for the correspondence analysis. The column on the left shows the variables as originally compiled. The column on the right shows how these have been aggregated for the analysis.", fig.height = 7}
# Most of the code for assembling the Sankey plot has been taken directly from 
# various walkthroughs at https://www.r-graph-gallery.com

@@

# Create plot
sn <- sankeyNetwork(Links = links, Nodes = nodes,
              Source = "IDsource", Target = "IDtarget",
              NodeID = "name", Value = "value", colourScale = my_colour,
              sinksRight = FALSE, LinkGroup = "new_categories", 
              NodeGroup="group", nodeWidth = 12, nodePadding = 7, fontSize = 10)

@@

```{r cor-plot, fig.cap = "Correlation matrix showing transformation of skewed variables for the PCA. The mean age of the sites has also been included to visualise overall temporal trends. Cells below the diagonal display the bivariate distributions with a fitted OLS-regression. The cells above the diagonal display and are coloured by the corresponding Pearson's correlation coefficient.", fig.height = 10.1, fig.width = 10.1}
# Transform the skewed variables - sqrt() is used where there are zero values,
# and seems to provide an adequate result.
lithicstr <- lithics2 %>%
  select(-site) %>% 
  # Transform skewed variables and reorder
  mutate(
         "log10(lithic_dens)" = log10(lithic_dens),
         "log10(secondary_prop)" = log10(secondary_prop),
         "core_prop^(1/3)" = core_prop^(1/3),
         "nonflint_prop^(1/3)" = nonflint_prop^(1/3), .keep = "unused") %>% 
  relocate("mean_date",
    "log10(lithic_dens)", "nonflint_prop^(1/3)", "blanks_prop",
    "log10(secondary_prop)", "chip_prop", "core_prop^(1/3)" 
  )

@@

The flint material of the later sites is to a larger extent characterised by micro-blades, which corresponds to the transition to micro-blade production from handle cores [e.g. @solheim2020]. A more fragmented flint material, as indicated by the relative importance of flint chips and fragments, is also a previously noted property of some later Mesolithic, as well as early Neolithic sites [e.g. @fossum2017; @stokke2018a]. The most defining material for the later sites, however, is non-flint macro tools and associated debitage, which is dominating some of these assemblages. It was noted above that this material does not seem to impact the latest sites, which would indicate that specialised axe production sites disappear towards the end of the Mesolithic, a notion that would be in line with previous suggestions [e.g. @glørstad2011; @reitan2016].

@@

@@

splt1 <- fviz_ca_row(nonprop, col.row = "black",
            fill = all_nonflint_prop$mean_date / 1000, title = "A",
                     repel = TRUE, shape = 21, pointsize = 2.9) +
  scale_fill_gradient2(name = "Mean age kyr BCE", low="yellow", mid="orange", 
                        high="red", midpoint = centdate) +
  ggtitle("Supplementary material") +
  theme(legend.position = "bottom",
        plot.title = element_text(size = 23, face = "bold",
                                  family = "serif"))

@@

```{r, fig.height = 15, fig.cap = "Same as above, only that here the proportion of non-flint is used as a supplementary column. The negligble difference between this CA and that above indicates that the flint/non-flint distinction is integrated in the different artefact types, and therefore that the effect of artefact types and raw material cannot be seperated. These plots thus hide important variability that is captured in the ones presented in the main text. A) Row plot, B) Column plot."}

# Version of the above where the proportion of non-flint is simply added as a 
# supplementary column, following suggestion from the reviewer. 
moddat <- select(all_nonflint_prop, -site, -mean_date, -flint_count, 
                       -reported_start_bc, -reported_end_bc,
                       -askeladden_id)

@@

splt4 <- fviz(nonpropsup, element = "col", col.col.sup = "black",
             color = "contrib", 
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE,
             title = "B") +
  labs(colour = "Contribution to total inertia") +
  theme(legend.position = "bottom")
